Common Tools and Community Software

### 2020-04-24 LHCC/2020/2020-01/ctcs.tex
Final version of text from Google Doc

% This is the styling for submission to arXiv using JHEP Preprint Style

@@

% JHEP preprint template
\documentclass[10pt,a4paper]{article}
\usepackage{jheppub}
\notoc

\usepackage{graphicx}
\usepackage{subcaption}
%\usepackage[titletoc]{appendix}
\usepackage{xspace}

@@

%% Comment out before final submission
\usepackage{lineno}  % for line numbering during review
\linenumbers

@@

Common and community software packages, such as ROOT, Geant4 and event
generators have been a key part of the LHC's success so far and
continued development and optimisation will be critical in the future
\textbackslash cite\{HSF-EPPSU-Document, EPPSU-Briefingbook\}. The
challenges are driven by an ambitious physics programme, notably the LHC
accelerator upgrade to high-luminosity, HL-LHC, and the corresponding
detector upgrades of ATLAS and CMS. The General Purpose Detectors
describe their specific challenges elsewhere; in this document we
address the issues for software that is used in multiple experiments
(usually even more widely than ATLAS and CMS) and maintained by teams of
developers who are either not linked to a particular experiment or who
contribute to common software within the context of their experiment
activity. We also give space to general considerations for future
software and projects that tackle upcoming challenges, no matter who
writes it, which is an area where community convergence on best practice
is extremely useful.

ATLAS and CMS will undergo major detector upgrades and will also
increase their trigger rates for HL-LHC by about a factor of 10; event
complexity will rise, with peak pile-up of 200, far higher than in
Run-2. This places an enormous burden on storage and processing
resources. Current CMOS microprocessor technology is clock speed limited
(due to the failure of Dennard scaling) and, while processors
per-integrated circuit still keeps rising, Moore's Law is expected to
stall during the 2020s. More importantly, the effective runtime related
improvements in computing from CPU servers at sites is likely to be only
in the 5-10\% per year range, making the shortfall in processing
resources more severe. As manufacturers struggle to deliver ever-more
effective computing through CPUs the drive to different architectures
intensifies, with GPUs becoming commonplace and increasing interest in
even more specialised architectures, such as TPUs, IPUs and developments
that make FPGA devices more user friendly. These pose huge challenges
for our community as the programming models and APIs vary widely here
and possible lock-in to a particular vendor's devices is a significant
problem for code sustainability and preservation. Huge work has been
done already to adapt to this changing landscape, e.g. multi-threading
software and GPU codes have been used in production by some experiments
for years. In other areas migration is still ongoing. In others it is
not even yet started. Generic heterogeneous programming models have
existed for some time, and new ones are arising, but there is, as yet,
no clear winner, in-part because the performance obtained can be highly
application-dependent. HEP itself will not drive the success of one
model or another, so even if the community converged on a prefered
model, its long term success would not be assured. The C++ standard
itself is likely to lag for many years behind what is required by us in
the area of heterogeneous or even distributed computing. Further,
experiment frameworks (and with knock-on effects to systems like
workload management) will need to adapt to sites that provide
heterogeneous resources. How to do this, and make effective use of
different heterogeneous resources across different sites, remains far
from settled.

@@

- {Career challenges}. Those working in the development, optimization or
execution of generator software provide essential contributions to the
success of the HL-LHC physics programme and it is critical that they get
the right recognition and motivation. However, theorists get recognition
on published papers, and may not be motivated to work on software
optimisations that are not "theoretical" enough to advance their
careers. Generator support tasks in the experiments may also not be
enough to secure jobs/funding to experimentalists pursuing a career in
research.

@@

- {Programming languages}: Attracting collaborators with a computer
science background to work on generators, especially students, may also
be complicated by the fact that critical components of some generator
packages are written in Fortran. In some cases, alternative
implementations of these components in other languages exist, but their
speed is significantly lower than the optimized Fortran version. Some of
the generators also do not use industry standard version control
systems, making it harder to contribute code.

@@

- {Merging inefficiency}. Merging prescriptions (e.g. MLM {[}Man02{]},
CKKW-L {[}Lon05{]} at LO and FxFx {[}Fre12{]}, MEPS@NLO {[}Hoe14{]} at
NLO) imply the rejection of some events, to avoid double counting
between events produced with n\textsubscript{jets}+1 MEs and with
n\textsubscript{jets} MEs plus parton showers. The resulting
inefficiencies can be relatively low depending on the process, but they
are unavoidable in the algorithmic strategy used by the underlying
physics modeling {[}Alw08{]}. However, a method like shower-kt MLM can
reduce the merging inefficiency of MLM {[}Alw09{]}.

@@

- {Negative weights}. Matching prescriptions, like MC@NLO, are required
in (N)NLO calculations to avoid double counting between (N)NLO matrix
elements and parton showers, leading to the appearance of events with
negative weights. This causes a large inefficiency, as larger event
samples must be generated and passed through the experiment simulation,
reconstruction and analysis codes, increasing the compute and storage
requirements. For a fraction r of negative weight events, the number of
events to generate increases by a factor 1/(1-2r)\^{}2: for instance,
with r=25\% (which may be regarded as a worst-case scenario in ttbar
production {[}Fre20{]}), one needs to generate 4 times as many events.
Negative weights can instead be almost completely avoided, by design, in
another popular matching prescription, POWHEG, which however is only
available for a limited number of processes and describes the relevant
physics to a different degree of precision than MC@NLO (see {[}Fre20{]}
for a more in-depth discussion). Progress in this area can only be
achieved by theorists: for instance, a new approach for MC@NLO-type
matching with reduced negative weights has recently been proposed
{[}Fre20{]}. It should be noted that negative weights can also happen at
LO because of not-positive-definite parton density function sets and
interference terms, e.g. in effective field theory calculations.

@@

3. {Size of unweighted event samples required for experimental analysis
at HL-LHC}. Another question is in which regions of phase space the
number of unweighted events must be strictly proportional to the
luminosity. For example, in the low pT regions of W boson production it
is probably impossible to keep up with the data, due to the huge cross
section. Alternative techniques should be investigated, to avoid the
generation of huge event samples.

\hypertarget{summary---generator-community-current-priorities}{%
\subsection{Summary - generator community current
priorities}\label{summary---generator-community-current-priorities}}

{[}Text moved to final summary{]}

@@

The Geant4 particle transport toolkit contains advanced models for
electromagnetic and hadronic physics processes, as well as the
geometrical description of the detectors. The toolkit design principles
were to allow flexibility in future extensions of its own code, as well
as choices for particular applications. It is clear that this
flexibility comes at the price of a performance penalty. We observe that
writing more compact and more specialised code can lead to several
percent speed-up {[}GeantV RD{]}. The technical improvements to explore
here consist of avoiding too many virtual layers and improving the
instruction and data locality, leading to better cache use. Going even
further as far as specialisation is concerned, studies suggest that
implementing compact, self-contained transport libraries with a minimal
set of physics models (for instance only electromagnetic interaction for
selected particles) with predefined scoring (to simulate, for instance,
the response of an EM calorimeter) can also bring speed-ups to the
complete application {[}GeantV RD{]}. As far as architectural
modifications are concerned, the design of the Geant4 track-status
machinery prevents the introduction of fine-grained track parallelism.
It is still not clear whether this kind of parallelism can bring any
substantial speed-up, due to the overheads; however, making the Geant4
kernel and physics processes stateless would certainly provide the
necessary starting points for further R\&D in that direction. Most of
these investigations have started and it is hoped to see some first
results on a year time-scale. More effort, however, will be required to
integrate any eventual changes in the production code in a timely
manner. The GeantV Vector prototype has allowed the assessment of the
achievable speedup using a novel, vectorized approach to particle
transport. The demonstrator of a full electromagnetic shower in
realistic (CMS) geometry has been implemented and the comparison to a
similar Geant4 application has been performed. The conclusion of that
work showed that the main factors in the speedup seem to include better
cache use and tighter code, while the vectorisation impact was much
smaller than hoped for. This unimpressive impact is due to a set of
factors, including the fraction of the current algorithms that could be
vectorised, unresolved challenges for the gather/scatter and tail
handling (to keep the number of events in flight within bound) needed
for large number of shapes, and of course Amdahl's Law applied to
vectorisation where some bottleneck (for example geometry navigation)
could not be tackled without a major additional long term effort.
{[}GeantV{]}. At the same time libraries developed in the context of the
GeantV R\&D, e.g. VecGeom, have been successfully integrated in Geant4
and ROOT benefitting the whole HEP software community {[}CMS01{]}. The
cost-over-benefit of more vectorisation of the code has not been deemed
worth the necessary investment also due to the speculation that a
complete rewrite of the physics base could become necessary to fully
exploit it. As a result investing in the optimisation and modernisation
of the Geant4 code has gained even more relevance.

@@

The so-called `fast simulation' techniques consist of replacing the
tracking of individual particles through (a part of) the detector,
including all the physics processes they would undergo, by a
parameterization where the detector response is produced directly as a
function of the incoming particle type and energy. An example of such a
parameterization was implemented several years ago in the context of the
H1 experiment. This parameterization, available within the Geant4
toolkit under the name of the GFLASH library, became the starting idea
for several `custom' implementations, specific for other calorimeters.
Those were interfaced to experiments' simulation applications through
the available `hook' in the Geant4 toolkit. The LHC experiments
implemented, for example, dedicated parametrized response libraries for
some of their calorimeters. Recently, an effort has been invested in
Geant4 to generalize the parameterization formulae, and to implement
automatic procedures of tuning the parameters for specific detector
geometries. This kind of `classical' fast simulation, which describes
the particle shower shape using some complex mathematical functions with
several parameters, will remain an important tool; however, it is also
clear that their overall precision for the different energy values,
especially for highly granular and complex calorimeters will always be
relatively low. The recent developments of Deep Learning-based
techniques have opened an exciting possibility of replacing those
`classical' parameterisations by trained neural networks that would
reproduce the detector response. This approach consists of training
generative models such as Generative Adversarial Networks (GAN),
Variational Auto-Encoders (VAE) or Autoregressive Generative Networks on
the `images' of particle showers {[}ML001{]}. The energy deposition
values in the calorimeter cells are considered as `pixels' of the images
that the network is supposed to reproduce. The first studies and
prototypes have shown very promising results, however, the
generalization of the developed models (for different calorimeters, with
different particles incident at different angles with different
energies), still requires further effort. Detailed physics validation of
those tools and understanding of the capability to reproduce the
fluctuations is a prerequisite towards `production-quality' fast
simulation libraries. This work is already ongoing, but will require,
over the next few years, more effort to be invested combining the
physics and Machine Learning expertise. Given that the training of the
deep learning algorithms usually requires using sufficiently large MC
samples produced using standard techniques, the development of the novel
tools does not necessarily remove the need to speed up Geant4.

It is worth pointing out that different fast simulation techniques have
already been extensively used for LHC simulation campaigns. Effort has
been spent in the experiments frameworks to combine as much as possible
fast and detailed implementation. In particular ATLAS and CMS are using
in their baseline production shower libraries for the modelling of
calorimeters in the forward region in combination with the detailed
simulation of the rest of the sub-detectors. In LHCb the re-use of part
of the underlying events for specific signals of interest has been
established in production of samples when appropriate, with particular
care paid to ensure keeping under control bias on statistical
uncertainties {[}LHCb02{]}, its applicability in ATLAS and CMS could be
explored for specific cases. Similarly, the re-use of simulated or real
events to mimic the background to hard-scatter events is also under
investigation, where additional challenges exist concerning storage and
I/O due to the handling of the large minimum bias samples needed to
model the additional interactions.

@@

@@

Implementations of real-time analysis systems, where raw data is
processed in its final form as close as possible to the detector (e.g.
in the high-level trigger farm), are in use within several experiments.
These approaches remove the detector data that typically makes up the
raw data tier kept for offline reconstruction, and keep only a limited
set of analysis objects reconstructed within the high level trigger.

>>>>

@@

% JHEP preprint template
\documentclass[10pt,a4paper]{article}
\usepackage{jheppub}
%\notoc

\usepackage{graphicx}
\usepackage{subcaption}
\usepackage[titletoc]{appendix}
\usepackage{xspace}

@@

% References
\usepackage[style=numeric-comp,sorting=none]{biblatex}
\addbibresource{ctcs.bib}

%% Comment out before final submission %%
\usepackage{lineno}  % for line numbering during review
\linenumbers

@@

Common and community software packages, such as ROOT, Geant4 and event
generators have been a key part of the LHC's success so far and
continued development and optimisation will be critical in the future
\cite{stewart_graeme_a_2018_2413005, Ellis:2691414}. The
challenges are driven by an ambitious physics programme, notably the LHC
accelerator upgrade to high-luminosity, HL-LHC, and the corresponding
detector upgrades of ATLAS and CMS. The General Purpose Detectors
describe their specific challenges elsewhere; in this document we
address the issues for software that is used in multiple experiments
(usually even more widely than ATLAS and CMS) and maintained by teams of
developers who are either not linked to a particular experiment or who
contribute to common software within the context of their experiment
activity. We also give space to general considerations for future
software and projects that tackle upcoming challenges, no matter who
writes it, which is an area where community convergence on best practice
is extremely useful.

ATLAS and CMS will undergo major detector upgrades and will also
increase their trigger rates for HL-LHC by about a factor of 10; event
complexity will rise, with peak pile-up of 200, far higher than in
Run-2. This places an enormous burden on storage and processing
resources. Current CMOS microprocessor technology is clock speed limited
(due to the failure of Dennard scaling) and, while processors
per-integrated circuit still keeps rising, Moore's Law is expected to
stall during the 2020s. More importantly, the effective runtime related
improvements in computing from CPU servers at sites is likely to be only
about 10\% per year, making the shortfall in processing resources more
severe. As manufacturers struggle to deliver ever-more effective
computing through CPUs the drive to different architectures intensifies,
with GPUs becoming commonplace and increasing interest in even more
specialised architectures, such as TPUs, IPUs and developments that make
FPGA devices more user friendly. These pose huge challenges for our
community as the programming models and APIs vary widely here and
possible lock-in to a particular vendor's devices is a significant
problem for code sustainability and preservation. Huge work has been
done already to adapt to this changing landscape, e.g. multi-threading
software and GPU codes have been used in production by some experiments
for years. In other areas migration is still ongoing. In others it is
not even yet started. Generic heterogeneous programming models have
existed for some time, and new ones are arising, but there is, as yet,
no clear winner, in-part because the performance obtained can be highly
application-dependent. HEP itself will not drive the success of one
model or another, so even if the community converged on a prefered
model, its long term success would not be assured. The C++ standard
itself is likely to lag for many years behind what is required by us in
the area of heterogeneous or even distributed computing. Further,
experiment frameworks (and with knock-on effects to systems like
workload management) will need to adapt to sites that provide
heterogeneous resources. How to do this, and make effective use of
different heterogeneous resources across different sites, remains far
from settled.

@@

- {Career challenges}. Those working in the development, optimization or
execution of generator software provide essential contributions to the
success of the (HL-)LHC physics programme and it is critical that they
get the right recognition and motivation. However, theorists get
recognition on published papers, and may not be motivated to work on
software optimisations that are not "theoretical" enough to advance
their careers. Generator support tasks in the experiments may also not
be enough to secure jobs/funding to experimentalists pursuing a career
in research.

@@

- {Programming languages}: Attracting collaborators with a computer
science background to work on generators, especially students, may also
be complicated by the fact that critical components of some generator
packages are written in Fortran, which is rarely used in industry and
less popular among developers than other programming languages. Some of
the generators also do not use industry standard version control
systems, making it harder to contribute code.

@@

- {Merging inefficiency}. Merging prescriptions (e.g. MLM {[}Man02{]},
CKKW-L {[}Lon05{]} at LO and FxFx {[}Fre12{]}, MEPS@NLO {[}Hoe14{]} at
NLO) imply the rejection of some events, to avoid double counting
between events produced with n\textsubscript{jets}+1 matrix elements and
with n\textsubscript{jets} MEs plus parton showers. The resulting
inefficiencies can be relatively high depending on the process, but they
are unavoidable in the algorithmic strategy used by the underlying
physics modeling {[}Alw08{]}. However, a method like shower-kt MLM can
reduce the merging inefficiency of MLM {[}Alw09{]}.

@@

- {Negative weights}. Matching prescriptions, like MC@NLO, are required
in (N)NLO calculations to avoid double counting between (N)NLO matrix
elements and parton showers, leading to the appearance of events with
negative weights. This causes a large inefficiency, as larger event
samples must be generated and passed through the experiment simulation,
reconstruction and analysis codes, increasing the compute and storage
requirements. For a fraction r of negative weight events, the number of
events to generate increases by a factor 1/(1-2r)\^{}2: for instance,
with r=25\% (which may be regarded as a worst-case scenario in top quark
pair production {[}Fre20{]}), one needs to generate 4 times as many
events. Negative weights can instead be almost completely avoided, by
design, in another popular matching prescription, POWHEG, which however
is only available for a limited number of processes and describes the
relevant physics to a different degree of precision than MC@NLO (see
{[}Fre20{]} for a more in-depth discussion). Progress in this area can
only be achieved by theorists: for instance, a new approach for
MC@NLO-type matching with reduced negative weights has recently been
proposed {[}Fre20{]}. It should be noted that negative weights can also
happen at LO because of not-positive-definite parton density function
sets and interference terms, e.g. in effective field theory
calculations.

@@

3. {Size of unweighted event samples required for experimental analysis
at HL-LHC}. Another question, unrelated to NNLO, is in which regions of
phase space the number of unweighted events must be strictly
proportional to the luminosity. For example, in the low pT regions of W
boson production it is probably impossible to keep up with the data, due
to the huge cross section. Alternative techniques should be
investigated, to avoid the generation of huge event samples.

\subsection{Generator References}\label{references}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% TO BE CONVERTED TO BibTeX, but left here for "working" %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

{[}Alb19{]} J. Albrecht et al. ``A Roadmap for HEP Software and
Computing R\&D for the 2020s'' Comput Softw Big Sci (2019) 3, 7,
\href{https://doi.org/10.1007/s41781-018-0018-8}{{doi:10.1007/s41781-018-0018-8}},
\href{https://arxiv.org/abs/1712.06982}{{arXiv:1712.06982}}

{[}Alw08{]} J. Alwall et al. ``Comparative study of various algorithms
for the merging of parton showers and matrix elements in hadronic
collisions'', Eur. Phys. J. C 53 (2008) 53,
\href{https://doi.org/10.1140/epjc/s10052-007-0490-5}{{doi:10.1140/epjc/s10052-007-0490-5}}

{[}Alw09{]} J. Alwall, S. de Visscher, F. Maltoni, ``QCD radiation in
the production of heavy colored particles at the LHC'', JHEP 0902 (2009)
017,
\href{https://doi.org/10.1088/1126-6708/2009/02/017}{{doi:10.1088/1126-6708/2009/02/017}}

{[}Alw14{]} J. Alwall et al. ``The automated computation of tree-level
and next-to-leading order differential cross sections, and their
matching to parton shower simulations'', JHEP 07 (2014) 079,
\href{https://doi.org/10.1007/JHEP07(2014)079}{{doi:10.1007/JHEP07(2014)079}}

{[}Bau13{]} C. Bauer et al., "Computing for perturbative QCD - a
Snowmass White Paper",
\href{https://arxiv.org/abs/1309.3598}{{arxiv:1309.3598}}

{[}Ben17{]} J. Bendavid, ``Efficient Monte Carlo Integration Using
Boosted Decision Trees and Generative Deep Neural Networks'',
\href{https://arxiv.org/abs/1707.00028}{{arXiv:1707.00028}}

{[}Bot19{]} E. Bothmann et al., ``Event generation with Sherpa 2.2'',
\href{https://scipost.org/SciPostPhys.7.3.034}{{SciPost Phys. 7 (2019)
034}}

{[}Bot20{]} E. Bothmann et al. ``Exploring phase space with Neural
Importance Sampling'',
\href{https://arxiv.org/abs/2001.05478}{{arXiv:2001.05478}}

{[}Chi17{]} J. T. Childers et al., "Adapting the serial Alpgen
parton-interaction generator to simulate LHC collisions on millions of
parallel threads", Comp. Phys. Comm. 210 (2017) 5154.
\href{https://doi.org/10.1016/j.cpc.2016.09.0135286}{{https://doi.org/10.1016/j.cpc.2016.09.0135286}}.

{[}Fre12{]} R. Frederix, S. Frixione, "Merging meets matching in
MC@NLO", JHEP 2012 (2012) 61.
\href{https://doi.org/10.1007/JHEP12(2012)061}{{doi:10.1007/JHEP12(2012)061}}

{[}Fre20{]} R. Frederix, S. Frixione, S. Prestel. P. Torrielli, ``On the
reduction of negative weights in MC@NLO-type matching procedures'',
\href{https://arxiv.org/abs/2002.12716}{{arXiv:2002.12716}}

{[}Fri17{]} S. Frixione, P. Nason, C. Oleari, "Matching NLO QCD
computations with parton shower simulations: the POWHEG method", JHEP11
(2007) 070,
\href{https://doi.org/10.1088/1126-6708/2007/11/070}{{doi:10.1088/1126-6708/2007/11/070}}

{[}Gao20{]} C. Gao. et al. ``Event Generation with Normalizing Flows'',
\href{https://arxiv.org/abs/2001.10028}{{arXiv:2001.10028}}

{[}Gen18{]} HSF Physics Event Generator Computing Workshop, CERN,
November 2018,
\href{https://indico.cern.ch/event/751693}{{https://indico.cern.ch/event/751693}}

{[}Gen19{]} HSF Physics Generators Working Group,
\href{https://hepsoftwarefoundation.org/workinggroups/generators.html}{{https://hepsoftwarefoundation.org/workinggroups/generators.html}}

{[}Gen20{]} HSF Physics Generators Working Group, "Challenges in Monte
Carlo event generator software for High-Luminosity LHC", in preparation
(2020).

{[}Gle08{]} T. Gleisberg, S. Hoeche, "Comix, a new matrix element
generator", JHEP 12 (2008) 039.
\href{https://doi.org/10.1088/1126-6708/2008/12/039}{{doi:10.1088/1126-6708/2008/12/039}}

{[}Hoe14{]} S. Hoeche, F. Krauss, M. Schoenherr, "Uncertainties in
MEPS@NLO calculations of h+jets", Phys. Rev. D 90 (2014) 014012.
\href{https://doi.org/10.1103/PhysRevD.90.014012}{{doi:10.1103/PhysRevD.90.014012}}

{[}Hoe19{]} S. Hoeche, S. Prestel, H. Schulz, "Simulation of vector
boson plus many final jets at the high luminosity LHC", Phys. Rev. D
100, 0140124 (2019).
\href{http://dx.doi.org/10.1103/PhysRevD.100.014024}{{doi:10.1103/PhysRevD.100.014024}}

{[}Jad03{]} S. Jadach, Foam: a general-purpose cellular Monte Carlo
event generator, Comp. Phys. Comm. 152 (2003) 55.
\href{https://doi.org/10.1016/S0010-4655(02)00755-5}{{doi:10.1016/S0010-4655(02)00755-5}}

{[}Kaw86{]} S.Kawabata, A new Monte Carlo event generator for high
energy physics, Comp. Phys. Comm. 41 (1986) 127.
\href{https://doi.org/10.1016/0010-4655(86)90025-1}{{doi:10.1016/0010-4655(86)90025-1}}

{[}Kli18{]} M.D. Klimek, M. Perelstein, ``Neural Network-Based Approach
to Phase Space Integration'',
\href{https://arxiv.org/abs/1810.11509}{{arXiv:1810.11509}}

{[}Kon20{]} D. Konstantinov, "Optimization of Pythia8", EP-SFT group
meeting, CERN, January 2020,
\href{https://indico.cern.ch/event/890670}{{https://indico.cern.ch/event/890670}}

{[}Lep80{]} G. P. Lepage, "VEGAS: an adaptive multi-dimensional
integration program", Cornell report CLNS-447 (1980).
\href{https://cds.cern.ch/record/123074}{{https://cds.cern.ch/record/123074}}

{[}Lon05{]} L. Lonnblad, "Correcting the colour-dipole cascade model
with fixed order matrix elements", JHEP 05 (2002) 046.
\href{https://doi.org/10.1088/1126-6708/2002/05/046}{{doi:10.1088/1126-6708/2002/05/046}}

{[}Mal03{]} F. Maltoni, T. Stelzer, "MadEvent: automatic event
generation with MadGraph", JHEP 02 (2003) 027.
\href{https://doi.org/10.1088/1126-6708/2003/02/027}{{doi:10.1088/1126-6708/2003/02/027}}

{[}Man02{]} M. L. Mangano, M. Moretti, R. Pittau, "Multijet matrix
elements and shower evolution in hadronic collisions: Wbb + n-jets as a
case study", Nucl. Phys. B 632 (2002) 343.
\href{https://doi.org/10.1016/S0550-3213(02)00249-3}{{doi:10.1016/S0550-3213(02)00249-3}}

{[}Mat16{]} O. Mattelaer, "On the maximal use of Monte Carlo samples:
re-weighting events at NLO accuracy", \emph{Eur. Phys. J. C}
76\textbf{,} 674 (2016),
\href{https://doi.org/10.1140/epjc/s10052-016-4533-7}{{doi:10.1140/epjc/s10052-016-4533-7}}

{[}Mat18{]} O. Mattelaer, "MG5aMC status and plans", in {[}Gen19{]}

{[}Nas07{]} P. Nason, MINT: a Computer Program for Adaptive Monte Carlo
Integration and Generation of Unweighted Distributions Bicocca-FT-07-13
(2007). \href{https://arxiv.org/abs/0709.2085}{{arxiv:0709.2085}}

{[}SKe18{]} L. Sexton-Kennedy, G. Stewart, "CWP challenges and workshop
aims", in {[}Gen19{]}

@@

The Geant4 particle transport toolkit contains advanced models for
electromagnetic and hadronic physics processes, as well as the
geometrical description of the detectors. The toolkit design principles
were to allow flexibility in future extensions of its own code, as well
as choices for particular applications. It is clear that this
flexibility comes at the price of a performance penalty. We observe that
writing more compact and more specialised code can lead in certain
places to up to several tens percent speed-up {[}GeantV RD{]}. The
technical improvements to explore here consist of avoiding too many
virtual layers and improving the instruction and data locality, leading
to better cache use. Going even further as far as specialisation is
concerned, studies suggest that implementing compact, self-contained
transport libraries with a minimal set of physics models (for instance
only electromagnetic interaction for selected particles) with predefined
scoring (to simulate, for instance, the response of an EM calorimeter)
can also bring speed-ups to the complete application {[}GeantV RD{]}. As
far as architectural modifications are concerned, the design of the
Geant4 track-status machinery prevents the introduction of fine-grained
track parallelism. It is still not clear whether this kind of
parallelism can bring any substantial speed-up, due to the overheads;
however, making the Geant4 kernel and physics processes stateless would
certainly provide the necessary starting points for further R\&D in that
direction. Most of these investigations have started and it is hoped to
see some first results on a year time-scale. More effort, however, will
be required to integrate any eventual changes in the production code in
a timely manner. The GeantV Vector prototype has allowed the assessment
of the achievable speedup using a novel, vectorized approach to particle
transport. The demonstrator of a full electromagnetic shower in
realistic (CMS) geometry has been implemented and the comparison to a
similar Geant4 application has been performed. The conclusion of that
work showed that the main factors in the speedup seem to include better
cache use and tighter code, while the vectorisation impact was much
smaller than hoped for. This unimpressive impact is due to a set of
factors, including the fraction of the current algorithms that could be
vectorised, unresolved challenges for the gather/scatter and tail
handling (to keep the number of events in flight within bound) needed
for large number of shapes, and of course Amdahl's Law applied to
vectorisation where some bottleneck (for example geometry navigation)
could not be tackled without a major additional long term effort.
{[}GeantV{]}. At the same time libraries developed in the context of the
GeantV R\&D, e.g. VecGeom, have been successfully integrated in Geant4
and ROOT benefitting the whole HEP software community {[}CMS01{]}. The
cost-over-benefit of more vectorisation of the code has not been deemed
worth the necessary investment also due to the speculation that a
complete rewrite of the physics base could become necessary to fully
exploit it. As a result investing in the optimisation and modernisation
of the Geant4 code has gained even more relevance.

@@

The so-called `fast simulation' techniques consist of replacing the
tracking of individual particles through (a part of) the detector,
including all the physics processes they would undergo, by a
parameterization where the detector response is produced directly as a
function of the incoming particle type and energy. An example of such a
parameterization was implemented many years ago in the context of the H1
experiment {[}H1Ref{]}. This parameterization, available within the
Geant4 toolkit under the name of the GFLASH library, became the starting
idea for several `custom' implementations, specific for other
calorimeters. Those were interfaced to experiments' simulation
applications through the available `hook' in the Geant4 toolkit. The LHC
experiments implemented, for example, dedicated parametrized response
libraries for some of their calorimeters. Recently, an effort has been
invested in Geant4 to generalize the parameterization formulae, and to
implement automatic procedures of tuning the parameters for specific
detector geometries. This kind of `classical' fast simulation, which
describes the particle shower shape using some complex mathematical
functions with several parameters, will remain an important tool;
however, it is also clear that their overall precision for the different
energy values, especially for highly granular and complex calorimeters
will always be relatively low. The recent developments of Deep
Learning-based techniques have opened an exciting possibility of
replacing those `classical' parameterisations by trained neural networks
that would reproduce the detector response. This approach consists of
training generative models such as Generative Adversarial Networks
(GAN), Variational Auto-Encoders (VAE) or Autoregressive Generative
Networks on the `images' of particle showers {[}ML001{]}. The energy
deposition values in the calorimeter cells are considered as `pixels' of
the images that the network is supposed to reproduce. The first studies
and prototypes have shown very promising results, however, the
generalization of the developed models (for different calorimeters, with
different particles incident at different angles with different
energies), still requires further effort. Detailed physics validation of
those tools and understanding of the capability to reproduce the
fluctuations is a prerequisite towards `production-quality' fast
simulation libraries. This work is already ongoing, but will require,
over the next few years, more effort to be invested combining the
physics and Machine Learning expertise. Given that the training of the
deep learning algorithms usually requires using sufficiently large MC
samples produced using standard techniques, the development of the novel
tools does not necessarily remove the need to speed up Geant4.

It is worth pointing out that different fast simulation techniques have
already been extensively used for LHC simulation campaigns. Effort has
been spent in the experiments frameworks to combine as much as possible
fast and detailed implementation. In particular ATLAS and CMS are using
in their baseline production shower libraries for the modelling of
calorimeters in the forward region in combination with the detailed
simulation of the rest of the sub-detectors. In LHCb the re-use of part
of the underlying events for specific signals of interest has been
established in production of samples when appropriate, with particular
care paid to ensure keeping under control bias on statistical
uncertainties {[}LHCb02{]}, its applicability in ATLAS and CMS could be
explored for specific cases. Similarly, the re-use of simulated or real
events to mimic the background to hard-scatter events is also exploited,
where additional challenges exist concerning storage and I/O due to the
handling of the large minimum bias samples needed to model the
additional interactions.

@@

\subsection{Simulation References}\label{references-1}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% TO BE CONVERTED TO BibTeX, but left here for "working" %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

{[}CMS01{]} V. Ivanchenko and S. Banerjee on behalf of the CMS
Collaboration, EPJ Web of Conferences 214, 02012 (2019), The PDF is
here:
\url{https://www.epj-conferences.org/articles/epjconf/pdf/2019/19/epjconf_chep2018_02012.pdf}

{[}CMS02{]} M Hildreth et al 2015 J. Phys.: Conf. Ser. 664 072022

{[}CMS03{]} M Hildreth et al 2017 J. Phys.: Conf. Ser. 898 042040

{[}GPU-``MPEXS-DNA{]} S. Okada et al. ``MPEXS-DNA: a new GPU-based Monte
Carlo simulator for track structures and radiation chemistry at
subcellular scale,'' Med Phys. vol 46 no. 3, pp. 1483-1500 (2019)

{[}Hybrid-Gate{]} J. Bert et al., "Hybrid GATE: A GPU/CPU implementation
for imaging and therapy applications," 2012 IEEE Nuclear Science
Symposium and Medical Imaging Conference Record (NSS/MIC), Anaheim, CA,
pp. 2247-2250 (2012)

{[}G4CU{]}
\url{https://kds.kek.jp/indico/event/15926/session/30/contribution/110/material/slides/0.pdf}

{[}Opticks{]} S. Blyth, ``Opticks : GPU Optical Photon Simulation for
Particle Physics using NVIDIA® OptiXTM,'' EPJ Web Conf., 214, pp. 02027
(2019) 02027

{[}LHCb01{]} D. Muller, ``Adopting new technologies in the LHCb Gauss
simulation framework'', EPJ Web of Conferences \textbf{214}, 02004
(2019); B.Siddi and D. Muller, ``Gaussino a Gaudi-based core simulation
framework'', CERN-LHCb-PROC-2019-010;

{[}ML001{]} Adversarial Networks (GAN) (3DGAN reference), Variational
Auto-Encoders (VAE) or Autoregressive Generative Networks on the
`images' of particle showers

{[}GAN-ATLAS, GAN-CMS, GAN-LHCb{]} Reference to GAN calorimeters for
ATLAS, CMS and LHCb

{[}LHCb02{]} D. Muller et al. ``ReDecay: A novel approach to speed up
the simulation at LHCb'' , Eur. Phys. J. C 78 (2018)

{[}GeantV RD{]}
\href{https://indico.cern.ch/event/818702/}{{https://indico.cern.ch/event/818702/}}

{[}CHEP-Juno{]} S. Blyth, ``Meeting the challenge of JUNO simulation
with Opticks GPU Optical Photon Accelerator'', presented at CHEP 2019 in
Adelaide, Australia

{[}SHIFT01{]} Hamilton, S.P., Evans, T.M., 2019. Continuous-energy Monte
Carlo neutron transport on GPUs in the Shift code. \emph{Annals of
Nuclear Energy} \textbf{128}, 236--247.
\url{https://doi.org/10.1016/j.anucene.2019.01.012}

{[}DELPHES{]} de Favereau, J., Delaere, C., Demin, P. \emph{et al.}
DELPHES 3: a modular framework for fast simulation of a generic collider
experiment. \emph{J. High Energ. Phys.} 2014, 57 (2014).
\href{https://doi.org/10.1007/JHEP02(2014)057}{{https://doi.org/10.1007/JHEP02(2014)057}}

{[}PGS4{]} J. Conway, R. Culbertson, R. Demina, B. Kilminster, M. Kruse,
S. Mrenna, J. Nielsen, M. Roco, A. Pierce, J. Thaler, T. Wizansky,
``Pretty Good Simulation'',
\href{http://conway.physics.ucdavis.edu/research/software/pgs/pgs4-general.htm}{{http://conway.physics.ucdavis.edu/research/software/pgs/pgs4-general.htm}}

{[}LAMARR{]} A.Davis, ``Fast simulations in LHCb'', CHEP 2019,
\href{https://indico.cern.ch/event/773049/contributions/3474742/attachments/1937708/3211775/davis_fast_simulations_at_lhcb_v3.pdf}{{https://indico.cern.ch/event/773049/contributions/3474742/attachments/1937708/3211775/davis\_fast\_simulations\_at\_lhcb\_v3.pdf}}

{[}ALICE01{]} B. Volker, HSF Simulation Working Group meeting,
\href{https://indico.cern.ch/event/782507/contributions/3293465/}{{https://indico.cern.ch/event/782507/contributions/3293465/}}

{[}H1Ref{]} G. Grindhammer, M. Rudowicz and S. Peters, NIM A290 (1990)
469-488

@@

Implementations of real-time analysis systems, where raw data is
processed into its final form as close as possible to the detector (e.g.
in the high-level trigger farm), are in use within several experiments.
These approaches remove the detector data that typically makes up the
raw data tier kept for offline reconstruction, and keep only a limited
set of analysis objects reconstructed within the high level trigger.
